Peptide retention time prediction

### 2014-11-08 retention.tex
Added a few paragraphs in the introduction

\section{Introduction}

Besides decreasing the number of peptides injected at the same time
into the mass spectrometer, the separation process in itself provides
useful information about the analytes. On the one hand, the instrument
always records the time when a peptide elutes from the chromatography
column, i.e. the retention time. On the other hand, the retention time
can also be estimated by examining the sequence of the peptide. The
availability of both predicted and observed retention times has proven
to be a powerful feature, with several applications in mass
spectrometry-based proteomics \cite{Palmblad2013}.  In this chapter,
Sections~\ref{sec:rtpred} and \ref{sec:rtpredm} describe the main
approaches used to estimate retention time from peptide sequences,
while Section~\ref{sec:app} discusses some of the most important
applications of retention time predictions. Section~\ref{sec:irt}
briefly introduces the use of peptide retention standards.

>>>>

\section{Introduction}
Mass spectrometry-based methods are unquestionably the most popular techniques for studying the proteint content of biological samples. The palette of protemic applications where such methods are used is getting increasingly diversified, ranging from identifying and quantifying
proteins in biological samples \cite{Geiger2012} to determining
posttranslational modifications \cite{Huttlin2010} and elucidating
protein interaction networks \cite{Gavin2011}.

Although available in a variety of flavours, most of the widely spread MS-based workflows consists of a few common steps. First, the proteins in the mixture of interest are cleaved into peptides using a sequence-specific enzyme such as trypsin. This step is necessary because the mass spectrometers are more sensitive for peptides, and because spectra produced by full-length proteins are difficult to interpret. In the next step, the resulting peptides are separated according to their hydrophobicity on a liquid chromatography (LC) column. As they gradually elute from the column, the peptides are ionized via electrospray ionization and subjected to mass spectrometry analysis. The result of these measurements consists of a set of spectra, which are subsequently used to infer valuable information about the proteins present in the initial sample.

Today a majority of the mass spectrometry (MS) protocols in use are high-throughput, generating extensive amounts of data. This has brought up a whole new set of analytical challenges, making the development of efficient algorithms to manipulate and interpret the data a major area of proteomics research. A particular difficult challenge in this context is posed by the tremendous complexity of the data. Proteins
display a wide range of physico-chemical properties and are often found at very different amounts in complex samples, spanning as much as 12 orders of magnitude in relative
abundance~\cite{Angel2012}. Furthermore, a large proportion of
proteins are subjected to posttranslational modifications. As an
example, a few years ago more than 25,000 phosphorylation sites on
human proteins were already reported~\cite{Lemeer2009}. To manage this formidable diversity,
 increasing efforts are made to extract useful information not only from the spectra, but also from other types of information recorded during MS experiments.

An example of such information is retention time, i.e. the time point when a peptide elutes from the liquid chromatography column. Although the main purpose of liquid chromatography is to reduce the number of peptides
simultaneously injected into the mass spectrometer, and by this
increase the proteome coverage achieved in an experiment. However, besides this, the separation process in itself provides
useful information about the analytes. On the one hand, the instrument
always records the time when a peptide elutes from the chromatography
column, i.e. the retention time. On the other hand, the retention time
can also be estimated by examining the sequence of the peptide. The
availability of both predicted and observed retention times has proven
to be a powerful feature, with several applications in mass
spectrometry-based proteomics \cite{Palmblad2013}.

In this review we summarize the work done in the field of retention time prediction, and describe a few of the successful applications of such predictions. 
Sections~\ref{sec:rtpred} and \ref{sec:rtpredm} describe the main
approaches used to estimate retention time from peptide sequences.
Section~\ref{sec:app} discusses some of the most important
applications of retention time predictions. Section~\ref{sec:irt}
briefly introduces the use of peptide retention standards.
